---
title: "Predicting Past Year Identity Theft with the 2016 Identity Theft Supplement of the National Crime Victimization Survey"
author: "Erika Harrell"
date: "August 2021"
output: 
  pdf_document: 
    fig_width: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Disclaimer: The points and views expressed in this document do not necessarily reflect the views of the Bureau of Justice Statistics nor that of the U.S. Department of Justice.

## Executive Summary
###  Data analysis was conducted to see if certain predictor variables were associated with past year identity theft. The data used in this analysis came from the 2016 Identity Theft Supplement (ITS) to the National Crime Victimization Survey (NCVS). Chi-square results found that identity theft was dependent on the majority of the predictors used (age, race/Hispanic origin, annual household income, being a victim of identity theft prior to the past year, having personal information exposed due to a data breach, and using preventative behaviors). Gender was found to be independent of past year identity theft (p> .05). Three logistic regression models were trained on the data, including one in which the outcome variable under went undersampling of cases where the respondent did not report past year identity theft. However, the best model was one without gender or preventative behaviors as predictors, with no undersampling of the outcome. This model had  higher precision and accuracy than the other 2 models.

```{r libraries}
#adding libraries
library(reshape2)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(caret)
library(caTools)
library(pROC)
```

## Loading Data
### Data were downloaded from the 2016 ITS study [page](https://www.icpsr.umich.edu/web/NACJD/studies/36829) from the National Archives of Criminal Justice Data to a subfolder entitled "data" in the project on R-Studio, unzipped, loaded into R-Studio and renamed with a shorter name. Due to the size of the datafile, the memory limit had to be increased prior to loading the data in R-Studio.

```{r load, echo=FALSE}
unzip("./data/ICPSR_36829-V2.zip", exdir="./data")
memory.limit(20000)
load(file="./data/ICPSR_36829/DS0001/36829-0001-Data.rda")
data<-da36829.0001
rm(da36829.0001)
```

## Data wrangling and cleaning
### Reducing the dataset- There were 125,165 total persons in the 2016 ITS dataset. Just using the completed  telephone and personal interviews leaves 96,130 interviews or observations in the dataset.

```{r interview,echo=FALSE}
x<-cbind(Count=table(data$VS008),Percentage=round(prop.table(table(data$VS008))*100))
rbind(x,Total = c(nrow(data),100))

data<-data[data$VS008 != levels(data$VS008)[3],]
x<-cbind(Count=table(data$VS008),Percentage=round(prop.table(table(data$VS008))*100))
rbind(x,Total = c(nrow(data),100))

rm(x)
```

### From the larger ITS dataset, several variables were used: past year identity theft (misuse or attempted misuse of an existing account, misuse or attempted misuse of personal inforamtion to opena new account of for some other fraudulent purpose) age, race/Hispanic origin, gender, annual household income, whether a respondet participated in  behaviors in the past year to prevent identity theft, whether a respondent experienced identity theft prior to the past year, whther a respondent was notified that their personal information was exposed due to a data breach.

```{r variable, echo=FALSE}
pastyearbankacct<-data$VS010
existing_bank<-data$VS012
currentccacct<-data$VS014
pastyearccacct<-data$VS016
existing_credit_card<-data$VS017
other_existing_accts<-data$VS019
open_new_acct<-data$VS041
personal_info<-data$VS063
sex<-data$V3018
race<-data$V3023A
hispanic<-data$V3024A
age<-data$V3014
income<-data$V2026A
OUTSIDE_PAST_YEAR<-data$VS306
CHCKD_CR_PAST_YR<-data$VS343
CHNG_PASSWORDS<-data$VS345
PURCHASE_IDTHFT_INS<-data$VS347
SHRED_DOCS<-data$VS349
VERIFY_CHARGES<-data$VS351
PROTECT_COMPUTER<-data$VS353
PURCHASE_IDTHFT_PROT<-data$VS355
notify_breach<-data$VS358
```

### Individual variables creaetd in the previous step were combined into a smaller dataset and larger dataset was removed.

```{r combine, echo=FALSE}
its<-cbind(data.frame(sex,race,hispanic,income, age,
                      pastyearbankacct, existing_bank, currentccacct, 
                      pastyearccacct, existing_credit_card, 
                      other_existing_accts, open_new_acct, personal_info,
                      OUTSIDE_PAST_YEAR, CHCKD_CR_PAST_YR, CHNG_PASSWORDS, 
                      PURCHASE_IDTHFT_INS, SHRED_DOCS,VERIFY_CHARGES, 
                      PROTECT_COMPUTER, PURCHASE_IDTHFT_PROT, 
                      notify_breach))
rm(sex,race,hispanic,income,age,
     pastyearbankacct, existing_bank, currentccacct, pastyearccacct, 
   existing_credit_card, other_existing_accts, open_new_acct,personal_info,
   OUTSIDE_PAST_YEAR,  CHCKD_CR_PAST_YR, CHNG_PASSWORDS, 
   PURCHASE_IDTHFT_INS, SHRED_DOCS, 
   VERIFY_CHARGES, PROTECT_COMPUTER, PURCHASE_IDTHFT_PROT, 
   notify_breach)
```

### Summary of the created dataset.

```{r summary, echo=FALSE}
str(its)
summary(its)
```


### Variables were recoded, collapsing categories and computing necessary categories and variables for analysis.

```{r recode,echo=FALSE}
its$idtheft<-"No past year ID theft"
its$idtheft[its$existing_bank=="(01) Yes"| its$existing_credit_card=="(01) Yes"| 
              its$other_existing_accts=="(01) Yes"| its$personal_info== "(01) Yes" |
              its$open_new_acct=="(01) Yes"]<-"Past year ID theft"
its$idtheft<-as.factor(its$idtheft)

its$sexr<-recode_factor(its$sex, '(1) Male'='Male', '(2) Female'='Female',
                        '(8) Residue'='Unknown')
its$sexr<-droplevels(its$sexr)

its$racer<-recode_factor(its$race, '(01) White only'='White',
      "(02) Black only"="Black",
      "(03) Am Ind/AK native only"='Other',
      "(04) Asian only"="Other",
      "(05) Hawaiian/Pacific IS only"="Other",
      "(06) White-Black"="2 or more races",
      "(07) White-Amer Ind"="2 or more races", 
      "(08) White-Asian"="2 or more races",
      "(09) White-Hawaiian"="2 or more races",
      "(10) Black-Amer Ind"="2 or more races",
      "(11) Black-Asian"="2 or more races",
      "(12) Black-Hawaiian/Pacific Ils"="2 or more races",
      "(13) American Indian-Asian"="Other",
      "(14) Asian-Hawaiian/Pacific Ils"="Other",
      "(15) White-Black-American Ind"="2 or more races",
      "(16) White-Black-Asian"="2 or more races",
      "(17) White-Amer Ind-Asian"="2 or more races",
      "(18) White-Asian-Hawaiian"="2 or more races",
      "(19) 2 or 3 races" ="2 or more races",
      "(20) 4 or 5 races"= "2 or more races",
      '(98) Residue' = 'Unknown') 

its$hispanicr<-recode_factor(its$hispanic, "(1) Yes"="Hispanic",  
                             "(2) No"="NonHispanic","(8) Residue"="Unknown")

its$ethnicr[its$racer=="White" & its$hispanicr=="NonHispanic"]<-"White"
its$ethnicr[its$racer=="Black" & its$hispanicr=="NonHispanic"]<-"Black"
its$ethnicr[its$hispanicr=="Hispanic"]<-"Hispanic"
its$ethnicr[its$racer=="Other" & its$hispanicr=="NonHispanic"]<-"Other race"
its$ethnicr[its$racer=="2 or more races" & its$hispanicr=="NonHispanic"]<-"2 or more races"
its$ethnicr<-as.factor(its$ethnicr)
its$ethnicr<-factor(its$ethnicr, levels = c("White","Black","Hispanic",
                                "Other race", "2 or more races"))

its$ager[its$age>=16 & its$age<=17]<-"Age 16 to 17"
its$ager[its$age>=18 & its$age<=24]<-"Age 18 to 24"
its$ager[its$age>=25 & its$age<=34]<-"Age 25 to 34"
its$ager[its$age>=35 & its$age<=49]<-"Age 35 to 49"
its$ager[its$age>=50 & its$age<=64]<-"Age 50 to 64"
its$ager[its$age>=65]<-"Age 65 or older"
its$ager<-as.factor(its$ager)

its$incomer<-recode_factor(its$income,  "(01) Less than $5,000"="$24,999 or less", 
    "(02) $5,000 to $7,499"="$24,999 or less",
    "(03) $7,500 to $9,999"="$24,999 or less",
    "(04) $10,000 to $12,499"="$24,999 or less",
    "(05) $12,500 to $14,999"="$24,999 or less",
    "(06) $15,000 to $17,499"="$24,999 or less",
    "(07) $17,500 to $19,999"="$24,999 or less",
    "(08) $20,000 to $24,999"="$24,999 or less",
    "(09) $25,000 to $29,999"="$25,000–$49,999",
    "(10) $30,000 to $34,999"="$25,000–$49,999", 
    "(11) $35,000 to $39,999"="$25,000–$49,999",
    "(12) $40,000 to $49,999"="$25,000–$49,999",
    "(13) $50,000 to $74,999"="$50,000–$74,999",
    "(14) $75,000 and over"="$75,000 or more")

its$OUTSIDE_PAST_YEARR<-recode_factor(its$OUTSIDE_PAST_YEAR,
                  "(02) No"="No prior to past year ID theft",
                  "(01) Yes" ="Prior to past year ID theft",
                  "(08) Residue"="Unknown",
                  "(98) Refused"="Unknown",
                  "(99) Don't know"="Unknown")
its$OUTSIDE_PAST_YEARR<-factor(its$OUTSIDE_PAST_YEARR,
                                levels = c("Unknown","No prior to past year ID theft",
                                           "Prior to past year ID theft")) 

its$prevent_total<-"Unknown"
its$prevent_total[its$CHCKD_CR_PAST_YR=="(01) Yes"|its$CHNG_PASSWORDS=="(01) Yes"|
                    its$PURCHASE_IDTHFT_INS=="(01) Yes"|its$SHRED_DOCS=="(01) Yes"|
                    its$VERIFY_CHARGES=="(01) Yes"| its$PROTECT_COMPUTER=="(01) Yes"|
                    its$PURCHASE_IDTHFT_PROT=="(01) Yes"]<-"At least one preventative behavior"
its$prevent_total[its$CHCKD_CR_PAST_YR=="(02) No" & its$CHNG_PASSWORDS=="(02) No"&
                    its$PURCHASE_IDTHFT_INS=="(02) No" & its$SHRED_DOCS=="(02) No"&
                    its$VERIFY_CHARGES=="(02) No" & its$PROTECT_COMPUTER=="(02) No"&
                    its$PURCHASE_IDTHFT_PROT=="(02) No"]<-"No preventative behaviors"
its$prevent_total<-as.factor(its$prevent_total)
its$prevent_total<-factor(its$prevent_total,levels = c("Unknown",
                                                       "No preventative behaviors",
                                                       "At least one preventative behavior"))

its$notify_breachr<-recode_factor(its$notify_breach,
                                  '(01) Yes'='Data breach victim',
                                  '(02) No'='Not a data breach victim',
                                  "(08) Residue"="Unknown",
                                  "(98) Refused"="Unknown",
                                  "(99) Don't know"="Unknown")
its$notify_breachr<-factor(its$notify_breachr,levels=c('Unknown','Not a data breach victim',
                                                       'Data breach victim'))
```

## Exploratory analysis
## Past year identity theft
### About 11% of the sample reported at least one type of identity theft (misuse of an existing account, misuse of personal information to open new account, or misuse of personal information for other fraudulent purposes) in the past year while 89% of the sample reported no identity theft in the past year.

```{r outcome,echo=FALSE}
x<-cbind(Count=table(its$idtheft),Percentage=round(prop.table(table(its$idtheft))*100))
rbind(x,Total = c(nrow(its),100))

ggplot(data = its, aes(x = idtheft, y=..prop.., group=1)) + 
  geom_bar(stat = "count", fill="lightgreen") +
  stat_count(geom = "text", colour = "black", size = 3.5,
             aes(label = paste0(round(..prop..*100,digits=2),"%")),
             position=position_stack(vjust=0.5)) +
  ggtitle("Identity theft in the past 12 months")+
  xlab("Past Year ID theft") + theme_clean()+
  ylab("Percent") + scale_y_continuous(labels=scales:: percent_format())
```

## Annual household income
### One in five (21%) respondents were in households with annual incomes of $24,999 or less. Another fourth (26%) of the sample were in households with annual incomes of $25,000 to $49,999. Eighteen percent (18%) of the sample were in households with annual incomes of $50,000 to $74,999. The reminder of the sample (35%) were in households with annual incomes of at least $75,000. Identity theft affected a higher percentage of those in households with higher annual incomes. Fifteen percent (15%) of persons in households with annual incomes of at least $75,000 reported experiencing identity theft in the past 12 months compared to 6.5% of those in household with annual incomes of $24,999 or less.

```{r income, echo=FALSE}
x<-cbind(Count=table(its$incomer),Percentage=round(prop.table(table(its$incomer))*100))
rbind(x,Total = c(nrow(its),100))

ggplot(its, aes(x=incomer, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count..,..x..,sum)[..x..]) ), stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Annual Household Income')+
  ggtitle("Annual Household Income by Identity Theft") +
  scale_fill_manual(values = c("blue", "lightblue"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Race/Hispanic origin
### Seventy-one percent (71%) of the sample were White while one in ten (10%) respondents were Black. Hispanics accounted for 13% of respondents. Persons who were of another race accounted for 5% of the sample. Persons of 2 or more races accounted for 1% of the sample. Identity theft appeared to account for a larger percentage of Whites (12%) and persons of 2 or more races (14%) than Blacks (8%), Hispanics (7%) and persons of other races (9%).

```{r race,echo=FALSE}
x<-cbind(Count=table(its$ethnicr),Percentage=round(prop.table(table(its$ethnicr))*100))
rbind(x,Total = c(nrow(its),100))

ggplot(its, aes(x=ethnicr, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count..,..x..,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Race/Hispanic origin')+
  ggtitle("Race/Hispanic Origin by Identity Theft") +
  scale_fill_manual(values = c("purple", "violet"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Age
### Twenty-eight(28%) of the sample was age 50 to 64 while nearly one in four (24%) were age 35 to 49. Twenty-three percent (23%) of the sample was age 65 or older. The remainder of the sample was under the age of 35. Thirteen percent (13%) of persons age 35 to 49 reported past year identity theft, compared to 7% of persons age 18 to 34 and 9% of persons age 65 or older.

```{r age, echo=FALSE}
x<-cbind(Count=table(its$ager),Percentage=round(prop.table(table(its$ager))*100))
rbind(x,Total = c(nrow(its),100))

ggplot(its, aes(x=ager, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Age')+ggtitle("Age by Identity Theft") +
  scale_fill_manual(values = c("darkred", "pink"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Gender
### More than half of the sample (53%) was female while the remainder (47%) was male. Past year identity theft was experienced by 11% of males and a similar percentage of females.

```{r gender, echo=FALSE}
x<-cbind(Count=table(its$sexr),Percentage=round(prop.table(table(its$sexr))*100))
rbind(x,Total = c(nrow(its),100))

ggplot(its, aes(x=sexr, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Gender')+ggtitle("Gender by Identity Theft") +
  scale_fill_manual(values = c("darkorange", "yellow"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Preventative behaviors
### Nearly nine out of ten respondents (88%) used at least one of the preventative behaviors measured (checked bank or credit card statements, shredded or destroyed documents with financial information, checked credit report, changed passwords on financial accounts, used identity-theft security program on computer, purchased identity theft insurance or credit monitoring service, purchased identity-theft protection) in the past 12 months to prevent being a victim of identity theft. Eighteen percent of persons who did not know if they had used a preventative behavior in the past 12 months reported past year identity theft compared to 12% of those who had used at least one preventative behavior in the past 12 months. Unusually, 2% of those who reported no preventative behaviors reported identity theft in the past year.

```{r prevent, echo=FALSE}
x<-cbind(Count=table(its$prevent_total),Percentage=round(prop.table(table(its$prevent_total))*100))
rbind(x,Total = c(nrow(its),100))

ggplot(its, aes(x=prevent_total, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Use of Preventative Behaviors')+
  ggtitle("Use At Least One Preventative Behaviors in the Past Year by Identity Theft") +
  scale_fill_manual(values = c("green", "lightgreen"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Identity theft prior to the past year
### Thirteen percent (13%) of the sample experienced identity theft (misuse of an existing account, misuse of personal information to create new account or misuse of personal information for other fraudulent purposes) prior to the 12 months prior to their ITS interview. The majority of the sample (87%) did not experience it. About one in five (19%) respondents who experienced identity theft prior to the past year also experienced identity theft in the past 12 months. About 10% of those who did not have identity theft prior to the past 12 months experienced identity theft in the past 12 months.

```{r prior, echo=FALSE}
x<-cbind(Count=table(its$OUTSIDE_PAST_YEARR),Percentage=round(prop.table(table(its$OUTSIDE_PAST_YEARR))*100))
rbind(x,Total = c(nrow(its),100))

ggplot(its, aes(x=OUTSIDE_PAST_YEARR, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('ID theft prior to the past year')+
  ggtitle("Identity Theft Prior to the Past Year by Past Year Identity Theft") +
  scale_fill_manual(values = c("brown", "tan"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Notified of personal information exposure due to data breach
### Eleven percent (11%) of the sample reported that they were notified that their personal information was exposed during a data breach. The majority of the sample (88%) reported that they were not notified that their personal information was exposed during a data breach. Of those who were notified that their information was exposed during a data breach, one in five (21%) reported being victims of identity theft in the past year. This is compared to 9% of those who were not notified that their personal information was exposed in a data breach being victims of past year identity theft.

```{r breach, echo=FALSE}
x<-cbind(Count=table(its$notify_breachr),Percentage=round(prop.table(table(its$notify_breachr))*100))
rbind(x,Total = c(nrow(its),100))

ggplot(its, aes(x=notify_breachr, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(color = "white",aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Notified of data breach in the past year')+
  ggtitle("Notified of Personal Information Exposed in Data Breach in the Past Year by Past Year Identity Theft") +
  scale_fill_manual(values = c("black", "darkgray"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
rm(x)
```

## Data analysis
## More data wrangling
###  Unknown level on each individual variable used in the analysis was changed to NA. Individual variables were combined into a single dataset. The number of complete (cases with no NA values on any variable) and incomplete cases (cases with at least one variable with a value of NA) was summed from the dataset. There were 614 cases with a value of NA on at least one variable with the remaining 95,516 cases being complete cases with no missing values.

```{r itsclean, echo=FALSE}
idtheft<-its$idtheft
incomer<-its$incomer
ager<-its$ager
ethnicr<-its$ethnicr
prevent_total<-its$prevent_total
OUTSIDE_PAST_YEARR<-its$OUTSIDE_PAST_YEARR
notify_breachr<-its$notify_breachr   
sexr<-its$sexr

levels(prevent_total)[levels(prevent_total)== "Unknown"]<-NA
levels(OUTSIDE_PAST_YEARR)[levels(OUTSIDE_PAST_YEARR)== "Unknown"]<-NA
levels(notify_breachr)[levels(notify_breachr)== "Unknown"]<-NA

its_clean<-cbind(data.frame(idtheft,incomer,ager,ethnicr,prevent_total,OUTSIDE_PAST_YEARR,notify_breachr,sexr))

incompletecases<-sum(!complete.cases(its_clean))
completecases<-sum(complete.cases(its_clean))
natable<-cbind(Number=c(incompletecases,completecases),Percent=round(prop.table(c(incompletecases,completecases))*100))
row.names(natable)<-c('Incomplete cases','Complete cases')
rbind(natable,Total=c(nrow(its_clean),100))

its_clean<-its_clean[complete.cases(its_clean),]

#check cleaned dataset with only complete cases
summary(its_clean)

rm(incompletecases,completecases,natable, 
   ager,ethnicr,idtheft,incomer,notify_breachr,OUTSIDE_PAST_YEARR,
   prevent_total,sexr)
```

## Correlation matrix
### Pearson correlations of the variables were generated. The predictors appear to be relatively independent of each other with no moderate or strong correlations in the dataset. Gender appeared to have no correlation to the outcome variable, past year identity theft (r=0).

```{r correlate,echo=FALSE}

its_clean_num<-its_clean
its_clean_num$sexr<-as.numeric(its_clean_num$sexr)-1
its_clean_num$ethnicr<-as.numeric(its_clean_num$ethnicr)-1
its_clean_num$ager<-as.numeric(its_clean_num$ager)-1
its_clean_num$incomer<-as.numeric(its_clean_num$incomer)-1
its_clean_num$idtheft<-as.numeric(its_clean_num$idtheft)-1
its_clean_num$OUTSIDE_PAST_YEARR<-as.numeric(its_clean_num$OUTSIDE_PAST_YEARR)-1
its_clean_num$prevent_total<-as.numeric(its_clean_num$prevent_total)-1
its_clean_num$notify_breachr<-as.numeric(its_clean_num$notify_breachr)-1


its_clean_cor<-round(cor(its_clean_num),2)
melted_its_clean_cor<-melt(its_clean_cor)
head(melted_its_clean_cor)
melted_its_clean_cor$Var1<-recode_factor(melted_its_clean_cor$Var1,
                                         "idtheft"="Past year ID theft",
                                         "incomer"="Annual household income",
                                         "ager"="Age",
                                         "ethnicr"="Race/Hispanic origin",
                                         "prevent_total"="Preventative behaviors",
                                         "OUTSIDE_PAST_YEARR"="ID theft prior to past year",
                                         "notify_breachr"="Data breach victim",
                                         "sexr"="Sex")
melted_its_clean_cor$Var2<-recode_factor(melted_its_clean_cor$Var2,
                                         "idtheft"="Past year ID theft",
                                         "incomer"="Annual household income",
                                         "ager"="Age",
                                         "ethnicr"="Race/Hispanic origin",
                                         "prevent_total"="Preventative behaviors",
                                         "OUTSIDE_PAST_YEARR"="ID theft prior to past year",
                                         "notify_breachr"="Data breach victim",
                                         "sexr"="Sex")

ggplot(data = melted_its_clean_cor, aes(x=Var1, y=Var2, fill=value)) + 
  ggtitle("Correlation Matrix of Variables")+geom_tile()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
  legend.title = element_blank(),axis.title=element_blank())+
  geom_text(aes(label=value), color="white")

rm(its_clean_cor,melted_its_clean_cor)
```

## Chi-Square Analysis
### Multiple chi-square analyses were run comparing each predictor to the outcome variable. The chi-square analyses show that between past year identity theft was dependent on most of the predictors (p < 0.05) with the exception of sex (p > 0.05).

```{r chi, echo=FALSE}
chisq.test(its_clean$idtheft,its_clean$incomer)
chisq.test(its_clean$idtheft,its_clean$ager)
chisq.test(its_clean$idtheft,its_clean$sexr)
chisq.test(its_clean$idtheft,its_clean$ethnicr)
chisq.test(its_clean$idtheft,its_clean$prevent_total)
chisq.test(its_clean$idtheft,its_clean$OUTSIDE_PAST_YEARR)
chisq.test(its_clean$idtheft,its_clean$notify_breachr)
```

## Machine learning 
### Models were trained onto the data in an attempt to predict past year identity theft. The data as split into a training and test dataset with 70% of cases in the train dataset and 30% of cases in the test dataset. A logistic regression model that contained all variables with the exception of gender (gender was deleted due to its lack of correlation with the outcome variable) was trained on the data. The resulting model showed that all of the predictors were statistically significant (p<.05) in predicting past year identity theft. Respondents who were in households with higher annual incomes, Whites and persons of 2 or more races, older persons,those who had experienced identity theft prior to the past 12 months, and those who were notified that there information was exposed in a data breach were more likely than others to report identity theft in the past year. One unusual finding in the model was that participation in at least one preventative measure was significant in being a victim of identity theft in the past year. This is addressed in the model below.The model was then assessed for accuracy, recall and precision. Its F1 score was also generated. The accuracy score was .58, meaning that the model only predicted 58% of cases correctly. Its precision was .56 meaning that when the model classified a respondent as a victim of identity theft in the past year, it was correct 56% of the time. Its recall was .94, meaning that it correctly idenitifed 94% of all victims of identity theft. Its F1 score was .70. Efforts were made to improve the model below.

```{r ml1, echo=FALSE}
set.seed(1234)
sampleSplit<-sample.split(its_clean$idtheft,SplitRatio=0.7)
trainSet<-its_clean[sampleSplit==TRUE,]
testSet<-its_clean[sampleSplit==FALSE,]
model<-glm(idtheft~incomer+ager+ethnicr+OUTSIDE_PAST_YEARR+prevent_total+notify_breachr
             ,data=trainSet,family="binomial")
summary(model)
probabs<-predict(model,testSet,type='response')
preds<-ifelse(probabs>0.11,1,0)
confusionMatrix(factor(preds),factor(as.numeric(testSet$idtheft)-1))
recall<-recall(factor(as.numeric(testSet$idtheft)-1),factor(preds))
precision<-precision(factor(as.numeric(testSet$idtheft)-1),factor(preds))
f1score<-2*((precision*recall)/(precision+recall))
print(paste0("The recall of the above model is ", recall,"."))
print(paste0("The precision of the above model is ", precision,"."))
print(paste0("The F1 score of the above model is ", f1score,"."))
```

### Another logistic regression model was trained on the data. Checking the 2016 ITS questionnaire revealed that questions surrounding preventative behaviors were somewhat vague and did not specify a specific temporal order for the preventative behaviors and the identity theft victimization. It is possible that a respondent could have engaged in the preventative behaviors after becoming a victim as well as prior to becoming a victim.  With this in mind, preventative behaviors was removed from the model and the model was trained on data that was again split (70% of cases in the train dataset/30% of cases in the test dataset) with a different seed set to protect against overfitting. In the model, again all predictors appeared to be statistically significant in predicting past year identity theft (p< .05) with similar results regarding comparisons as the previous model. The model was then used to predict the outcome in the test data and metrics used to assess the quality of the model were generated. The accuracy score (.64), F1 score (.78), and  the precision (.65) improved while the recall (.93) declined. Another improvement to the data was made and assessed below.

```{r ml2, echo=FALSE}
set.seed(123)
sampleSplit<-sample.split(its_clean$idtheft,SplitRatio=0.7)
trainSet<-its_clean[sampleSplit==TRUE,]
testSet<-its_clean[sampleSplit==FALSE,]
model<-glm(idtheft~incomer+ager+ethnicr+OUTSIDE_PAST_YEARR+notify_breachr
           ,data=trainSet,family="binomial")
summary(model)
probabs<-predict(model,testSet,type='response')
preds<-ifelse(probabs>0.11,1,0)
confusionMatrix(factor(preds),factor(as.numeric(testSet$idtheft)-1))
recall<-recall(factor(as.numeric(testSet$idtheft)-1),factor(preds))
precision<-precision(factor(as.numeric(testSet$idtheft)-1),factor(preds))
f1score<-2*((precision*recall)/(precision+recall))
print(paste0("The recall of the above model is ", recall,"."))
print(paste0("The precision of the above model is ", precision,"."))
print(paste0("The F1 score of the above model is ", f1score,"."))
```

### In an attempt to improve the model used to predict past year identity theft, changes to the data itself were made. Since the data is imbalanced in terms of the outcome variable (89% with no past year identity theft vs 11% reporting identity theft in the past year), random undersampling of the cases where there was no past year identity theft, and keeping all of the cases in which the respondent reported identity theft in place. This resulted in a dataset with an outcome variable that had the same number of cases in which the respondent reported identity theft and respondents reported no past year identity theft (10,258). While the precision remained the same as the previous model, the accuracy score went down (.61), the F1 score (.62), recall (.60), and precision (.64) all decreased compared to the previous model, suggesting that undersampling or some other adjustment to the outcome variable might not be necessary for ITS data with its imbalanced outcome. Therefore, the second logistic regression model is suggested to predict past year identity theft.

```{r undersample,echo=FALSE}
theft<-its_clean[its_clean$idtheft=="Past year ID theft",]
notheft1<-its_clean[its_clean$idtheft!="Past year ID theft",]
set.seed(12)
notheft<-notheft1[sample(nrow(notheft1),nrow(theft)),]
newits<-rbind(theft,notheft)
table(newits$idtheft)
set.seed(12)
sampleSplit<-sample.split(newits$idtheft,SplitRatio=0.7)
trainSet<-newits[sampleSplit==TRUE,]
testSet<-newits[sampleSplit==FALSE,]
model<-glm(idtheft~incomer+ager+ethnicr+OUTSIDE_PAST_YEARR+notify_breachr
           ,data=trainSet,family="binomial")
summary(model)
probabs<-predict(model,testSet,type='response')
preds<-ifelse(probabs>0.5,1,0)
confusionMatrix(factor(preds),factor(as.numeric(testSet$idtheft)-1))
recall<-recall(factor(as.numeric(testSet$idtheft)-1),factor(preds))
precision<-precision(factor(as.numeric(testSet$idtheft)-1),factor(preds))
f1score<-2*((precision*recall)/(precision+recall))
print(paste0("The recall of the above model is ", recall,"."))
print(paste0("The precision of the above model is ", precision,"."))
print(paste0("The F1 score of the above model is ", f1score,"."))
```

