---
title: "Predicting Past Year Identity Theft with the 2016 Identity Theft Supplement of the National Crime Victimization Survey"
author: "Erika Harrell"
date: "12/21/2020"
output: 
  pdf_document: 
    fig_width: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
###  Data analysis was conducted to see if certain predictor variables were associated with past year identity theft. The data used in this analysis came from the 2016 Identity Theft Supplement (ITS) to the National Crime Victimization Survey (NCVS). The results of this analysis found that identity theft was dependent on the majority of the predictors used (age, race/Hispanic origin, annual household income, being a victim of identity theft prior to the past year, having personal information exposed due to a data breach, and using preventative behaviors). Gender was found to be independent of past year identity theft.A logistic regression model, with all predictors excluding sex, predicting past year identity theft was trained on the data. All predictors were found to be significant in predicting the outcome.

```{r libraries}
#adding libraries
library(dplyr)
library(ggplot2)
library(ggthemes)
library(caret)
library(car)
```

## Loading Data
### Data were downloaded from the 2016 ITS study [page](https://www.icpsr.umich.edu/web/NACJD/studies/36829) from the National Archives of Criminal Justice Data to a subfolder entitled "data" in the project on R-Studio, unzipped, loaded into R-Studio and renamed with a shorter name. Due to the size of the datafile, the memory limit had to be increased prior to loading the data in R-Studio.

```{r load, echo=FALSE}
unzip("./data/ICPSR_36829-V1.zip", exdir="./data")
memory.limit(20000)
load(file="./data/ICPSR_36829/DS0001/36829-0001-Data.rda")
data<-da36829.0001
rm(da36829.0001)
```

## Data Wrangling
### There were 125,165 total persons in the 2016 ITS. Only completed telephone and personal interviews were used in the analysis which left 96,130 interviews or observations in the dataset.

```{r interview,echo=FALSE}
Number <- table(data$VS008)
Number <- cbind(Number, Total = nrow(data))
Number <- rbind(Number, Total = nrow(data))
Number<-subset(Number,select=-Total)
Number

data<-data[data$VS008 != levels(data$VS008)[3],]
Number <- table(data$VS008)
Number <- cbind(Number, Total = nrow(data))
Number <- rbind(Number, Total = nrow(data))
Number<-subset(Number,select=-Total)
Number
rm(Number)
```

### Created variables that would be used in analysis from the larger ITS dataset.

```{r variable, echo=FALSE}
sex<-data$V3018
race<-data$V3023A
hispanic<-data$V3024A
age<-data$V3014
income<-data$V2026A
pastyearbankacct<-data$VS010
existing_bank<-data$VS012
currentccacct<-data$VS014
pastyearccacct<-data$VS016
existing_credit_card<-data$VS017
other_existing_accts<-data$VS019
open_new_acct<-data$VS041
personal_info<-data$VS063
OUTSIDE_PAST_YEAR<-data$VS306
CHCKD_CR_PAST_YR<-data$VS343
CHNG_PASSWORDS<-data$VS345
PURCHASE_IDTHFT_INS<-data$VS347
SHRED_DOCS<-data$VS349
VERIFY_CHARGES<-data$VS351
PROTECT_COMPUTER<-data$VS353
PURCHASE_IDTHFT_PROT<-data$VS355
notify_breach<-data$VS358
```

### Individual variables creatd in the previous step were combined into a smaller dataset and larger dataset was removed.

```{r combine, echo=FALSE}

its<-cbind(data.frame(sex,race,hispanic,income, age,
                      pastyearbankacct, existing_bank, currentccacct, 
                      pastyearccacct, existing_credit_card, 
                      other_existing_accts, open_new_acct, personal_info,
                      OUTSIDE_PAST_YEAR, 
                      CHCKD_CR_PAST_YR, CHNG_PASSWORDS, PURCHASE_IDTHFT_INS, SHRED_DOCS,  
                      VERIFY_CHARGES, PROTECT_COMPUTER, PURCHASE_IDTHFT_PROT, 
                      notify_breach
))


rm(sex,race,hispanic,income,age,
     pastyearbankacct, existing_bank, currentccacct, pastyearccacct, 
   existing_credit_card, other_existing_accts, open_new_acct,personal_info,
   OUTSIDE_PAST_YEAR, 
   CHCKD_CR_PAST_YR, CHNG_PASSWORDS,  PURCHASE_IDTHFT_INS, SHRED_DOCS, 
   VERIFY_CHARGES, PROTECT_COMPUTER, PURCHASE_IDTHFT_PROT, 
   notify_breach, data)
```

### Summary of the created dataset.

```{r summary, echo=FALSE}
str(its)
summary(its)
```


### Recoded variables, collapsing categories and computing necessary categories and variables for analysis.

```{r recode,echo=FALSE}

its$sexr<-recode_factor(its$sex, '(1) Male'='Male', '(2) Female'='Female', '(8) Residue'='Unknown')

its$racer<-recode_factor(its$race, '(01) White only'='White',
                         "(02) Black only"="NonWhite","(03) Am Ind/AK native only"='NonWhite',"(04) Asian only"='NonWhite',
                         "(05) Hawaiian/Pacific IS only"='NonWhite',"(06) White-Black"="NonWhite"  , "(07) White-Amer Ind"="NonWhite",      
                         "(08) White-Asian"="NonWhite","(09) White-Hawaiian"="NonWhite","(10) Black-Amer Ind"="NonWhite",       
                         "(11) Black-Asian"="NonWhite","(12) Black-Hawaiian/Pacific Ils"="NonWhite","(13) American Indian-Asian"="NonWhite",
                         "(14) Asian-Hawaiian/Pacific Ils"="NonWhite",
                         "(15) White-Black-American Ind"="NonWhite","(16) White-Black-Asian"="NonWhite","(17) White-Amer Ind-Asian"="NonWhite",
                         "(18) White-Asian-Hawaiian"="NonWhite","(19) 2 or 3 races" ="NonWhite", "(20) 4 or 5 races"= "NonWhite",'(98) Residue' = 'Unknown')           


its$hispanicr<-recode_factor(its$hispanic, "(1) Yes"="Hispanic",  "(2) No"="NonHispanic","(8) Residue"="Unknown")


its$ethnicr[its$racer=="White" & its$hispanicr=="NonHispanic"]<-"NonHispanic White"
its$ethnicr[its$racer!="White" | its$hispanicr!="NonHispanic"]<-"Not NonHispanic White"
its$ethnicr<-as.factor(its$ethnicr)

its$ager[its$age>=35 & its$age<=49]<-"Age 35 to 49"
its$ager[its$age>=16 & its$age<=17]<-"Age 16 to 17"
its$ager[its$age>=18 & its$age<=24]<-"Age 18 to 24"
its$ager[its$age>=25 & its$age<=34]<-"Age 25 to 34"
its$ager[its$age>=50 & its$age<=64]<-"Age 50 to 64"
its$ager[its$age>=65]<-"Age 65 or older"
its$ager<-as.factor(its$ager)

its$incomer<-recode_factor(its$income,  "(01) Less than $5,000"="Less than $75,000", "(02) $5,000 to $7,499"="Less than $75,000","(03) $7,500 to $9,999"="Less than $75,000" ,
                           "(04) $10,000 to $12,499"="Less than $75,000" ,"(05) $12,500 to $14,999"="Less than $75,000","(06) $15,000 to $17,499"="Less than $75,000",
                           "(07) $17,500 to $19,999"="Less than $75,000", "(08) $20,000 to $24,999"="Less than $75,000","(09) $25,000 to $29,999"="Less than $75,000",
                           "(10) $30,000 to $34,999"="Less than $75,000", "(11) $35,000 to $39,999"="Less than $75,000","(12) $40,000 to $49,999"="Less than $75,000",
                           "(13) $50,000 to $74,999"="Less than $75,000","(14) $75,000 and over"="$75,000 or more")

its$idtheft<-"No past year ID theft "
its$idtheft[its$existing_bank=="(01) Yes"| its$existing_credit_card=="(01) Yes"| 
              its$other_existing_accts=="(01) Yes"| its$personal_info== "(01) Yes" |
              its$open_new_acct=="(01) Yes"]<-"Past year ID theft"
its$idtheft<-as.factor(its$idtheft)

its$OUTSIDE_PAST_YEARR[its$OUTSIDE_PAST_YEAR=="(02) No"]<-"No prior to past year ID theft"
its$OUTSIDE_PAST_YEARR[its$OUTSIDE_PAST_YEAR=="(01) Yes" ]<-"Prior to past year ID theft"
its$OUTSIDE_PAST_YEARR[its$OUTSIDE_PAST_YEAR=="(08) Residue"|
                         its$OUTSIDE_PAST_YEAR== "(98) Refused"|
                         its$OUTSIDE_PAST_YEAR=="(99) Don't know"]<-"Unknown"
its$OUTSIDE_PAST_YEARR<-as.factor(its$OUTSIDE_PAST_YEARR)

its$prevent_total<-"Unknown"
its$prevent_total[its$CHCKD_CR_PAST_YR=="(01) Yes"|its$CHNG_PASSWORDS=="(01) Yes"|
                    its$PURCHASE_IDTHFT_INS=="(01) Yes"|its$SHRED_DOCS=="(01) Yes"|
                    its$VERIFY_CHARGES=="(01) Yes"| its$PROTECT_COMPUTER=="(01) Yes"|
                    its$PURCHASE_IDTHFT_PROT=="(01) Yes"]<-"Yes"
its$prevent_total[its$CHCKD_CR_PAST_YR=="(02) No" & its$CHNG_PASSWORDS=="(02) No"&
                    its$PURCHASE_IDTHFT_INS=="(02) No" & its$SHRED_DOCS=="(02) No"&
                    its$VERIFY_CHARGES=="(02) No" & its$PROTECT_COMPUTER=="(02) No"&
                    its$PURCHASE_IDTHFT_PROT=="(02) No"]<-"No"
its$prevent_total<-as.factor(its$prevent_total)

its$notify_breachr<-"Unknown"
its$notify_breachr[its$notify_breach=='(01) Yes']<-'Data breach victim'
its$notify_breachr[its$notify_breach=='(02) No']<-'Not a data breach victim'
its$notify_breachr[its$notify_breach=="(08) Residue"|its$notify_breach=="(98) Refused"|its$notify_breach=="(99) Don't know"]<-"Unknown"
its$notify_breachr<-as.factor(its$notify_breachr)
```

## Exploratory analysis
## Past year identity theft
### About 11% of the sample reported at least one type of identity theft (misuse of an existing account, misuse of personal information to open new account, or misuse of personal information for other fraudulent purposes) in the past year while 89% of the sample reported no identity theft.

```{r outcome,echo=FALSE}
ggplot(data = its, aes(x = idtheft, y=..prop.., group=1)) + 
  geom_bar(stat = "count", fill="lightgreen") +
  stat_count(geom = "text", colour = "black", size = 3.5,
             aes(label = paste0(round(..prop..*100,digits=2),"%")),
             position=position_stack(vjust=0.5)) +
  ggtitle("Identity theft in the past 12 months")+
  xlab("Past Year ID theft") + theme_clean()+
  ylab("Percent") + scale_y_continuous(labels=scales:: percent_format())
```

## Annual household income
### About two thirds of the sample were in households with annual incomes of less than $75,000 (65%) while the remainder (35%) were in households with annual incomes of at least $75,000. Within each income category the majority of respondents did not report experiencing identity theft in the past year. However, 15% of persons in households with incomes of $75,000 or more reported past year identity theft, compared to 9% of those in other households.

```{r income, echo=FALSE}
percentage<-prop.table(table(its$incomer))
x<-cbind(Count=table(its$incomer), Percentage=round(percentage*100))
rbind(Total = c(nrow(its),100),x)

ggplot(its, aes(x=incomer, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count..,..x..,sum)[..x..]) ), stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Annual Household Income')+
  ggtitle("Annual Household Income by Identity Theft") +
  scale_fill_manual(values = c("blue", "lightblue"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Race/Hispanic origin
### Seventy one percent of respondents were NonHispanic White while 29% were belonged to other race/Hispanic origin groups. Twelve percent of nonHispanic Whites reported past year identity theft compared to 8% of other persons reported identity theft in the past year.

```{r race,echo=FALSE}
percentage<-prop.table(table(its$ethnicr))
x<-cbind(Count=table(its$ethnicr), Percentage=round(percentage*100))
rbind(Total = c(nrow(its),100),x)

ggplot(its, aes(x=ethnicr, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count..,..x..,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Race/Hispanic origin')+
  ggtitle("Race/Hispanic Origin by Identity Theft") +
  scale_fill_manual(values = c("purple", "violet"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Age
### Twenty eight percent of the sample was age 50 to 64 while nearly one in four (24%) were age 35 to 49. 23% of the sample was age 65 or older. The remainder of the sample was under the age of 35. Thirteen percent of persons age 35 to 49 reported experiencing past year identity theft, compared to 7% of persons age 18 to 34 and 9% of persons age 65 or older.

```{r age, echo=FALSE}
percentage<-prop.table(table(its$ager))
x<-cbind(Count=table(its$ager), Percentage=round(percentage*100))
rbind(Total = c(nrow(its),100),x)

ggplot(its, aes(x=ager, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Age')+ggtitle("Age by Identity Theft") +
  scale_fill_manual(values = c("darkred", "pink"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Gender
### More than half of the sample (53%) was female while the remainder (47%) was male. Past year identity theft was experienced by 11% of males and a similar percentage of females.

```{r gender, echo=FALSE}
percentage<-prop.table(table(its$sexr))
x<-cbind(Count=table(its$sexr), Percentage=round(percentage*100))
rbind(Total = c(nrow(its),100),x)

ggplot(its, aes(x=sexr, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Gender')+ggtitle("Gender by Identity Theft") +
  scale_fill_manual(values = c("darkorange", "yellow"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Preventative behaviors
### Nearly nine out of ten persons in the sample (88%) used at least one of the preventative behaviors measured (checked credit report, changed password on financial accounts, had credit monitoring services or identity theft insurance, shredded or destroyed documents containing personal information, checked bank or credit card statements for unfamiliar charges, purchased identity theft protection) in the past 12 months. Eighteen percent of respondents who did not know if they had used a preventative behavior in the past 12 months reported past year identity theft compared to 12% of those who had used at least 1 preventative behavior in the past 12 months. Surprisingly, only 2% of those who did not participate in any prventative behaviors in the past 12 months reported past year identity theft.

```{r prevent, echo=FALSE}
percentage<-prop.table(table(its$prevent_total))
x<-cbind(Count=table(its$prevent_total), Percentage=round(percentage*100))
rbind(Total = c(nrow(its),100),x)

ggplot(its, aes(x=prevent_total, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Use of Preventative Behaviors')+
  ggtitle("Use At Least One Preventative Behaviors in the Past Year by Identity Theft") +
  scale_fill_manual(values = c("green", "lightgreen"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Identity theft prior to the past year
### Thirteen percent of the sample experienced identity theft (misuse of an existing account, misuse of personal information to create new account or misuse of personal information for other fraudulent purposes) prior to 12 months prior to their ITS interview. The majority of the sample did not experience it. About 40% of respondents who did not know if they were a victim of identity theft prior to the past year experienced identity theft in the past 12 months. This is compared to 10% of those who had no identity theft prior to the past year and 19% of those who had identity theft prior to the past year.

```{r prior, echo=FALSE}
percentage<-prop.table(table(its$OUTSIDE_PAST_YEARR))
x<-cbind(Count=table(its$OUTSIDE_PAST_YEARR), Percentage=round(percentage*100))
rbind(Total = c(nrow(its),100),x)

ggplot(its, aes(x=OUTSIDE_PAST_YEARR, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('ID theft prior to the past year')+
  ggtitle("Identity Theft Prior to the Past Year by Past Year Identity Theft") +
  scale_fill_manual(values = c("brown", "tan"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Notified of exposure due to data breach
### Twelve percent of the sample reported that they were notified that their personal information was exposed during a data breach. The majority of the sample (88%) reported that they were not notified that their personal information was exposed during a data breach.Of those who were notified that their information was exposed during a data breach, one in five (21%) reported being victims of identity theft in the past year. This is compared to 9% of those who were not notified that their personal information was exposed in a data breach being victims of past year identity theft.

```{r breach, echo=FALSE}
percentage<-prop.table(table(its$notify_breachr))
x<-cbind(Count=table(its$notify_breachr), Percentage=round(percentage*100))
rbind(Total = c(nrow(its),100),x)

ggplot(its, aes(x=notify_breachr, fill=idtheft))+
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..]), position="stack") +
  geom_text(color = "white",aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], 
                 label=scales::percent(..count../tapply(..count.., ..x.. ,sum)[..x..]) ),
            stat="count", position=position_dodge(0), vjust=1.5)+
  ylab('Percent') + xlab('Notified of data breach in the past year')+
  ggtitle("Notified of Personal Information Exposed in Data Breach in the Past Year by Past Year Identity Theft") +
  scale_fill_manual(values = c("black", "darkgray"))+
  scale_y_continuous(labels = scales::percent)+
  theme_clean()+labs(fill = "Past Year Identity Theft")
```

## Data analysis
## More data wrangling
### Make copies of each variable used in analysis. Unknown level on each individual variable was changed to NA. Individual  variables were combined into a single dataset and deleted individual variables. The individual variables were combined into a dataset (its1). The number of complete (cases with no NA values on any variable) and incomplete cases (cases with at least one variable with a value of NA) was summed from the dataset. There were 614 cases with a value of NA on atleast one variable with the remaining cases being complete cases

```{r its1, echo=FALSE}
idtheft<-its$idtheft
incomer<-its$incomer
ager<-its$ager
ethnicr<-its$ethnicr
prevent_total<-its$prevent_total
OUTSIDE_PAST_YEARR<-its$OUTSIDE_PAST_YEARR
notify_breachr<-its$notify_breachr
sexr<-its$sexr

levels(idtheft)[levels(idtheft)== "Unknown"]<-NA
levels(incomer)[levels(incomer)== "Unknown"]<-NA
levels(ager)[levels(ager)== "Unknown"]<-NA
levels(ethnicr)[levels(ethnicr)== "Unknown"]<-NA
levels(prevent_total)[levels(prevent_total)== "Unknown"]<-NA
levels(OUTSIDE_PAST_YEARR)[levels(OUTSIDE_PAST_YEARR)== "Unknown"]<-NA
levels(notify_breachr)[levels(notify_breachr)== "Unknown"]<-NA
levels(sexr)[levels(sexr)== "Unknown"]<-NA

its1<-cbind(data.frame(idtheft,incomer,ager,ethnicr,prevent_total,OUTSIDE_PAST_YEARR,notify_breachr,sexr))

str(its1)

incompletecases<-sum(!complete.cases(its1))
completecases<-sum(complete.cases(its1))
x<-c(incompletecases,completecases)
percentage<-round(prop.table(x)*100)
natable<-cbind(Number=x,Percent=percentage)
row.names(natable)<-c('Cases with NAs','Cases without NAs')
rbind(Total=c(nrow(its1),100),natable)
rm(incompletecases,completecases,x,percentage,natable)

```

## Data analysis
## Chi-Square Analysis
### Multiple chi-square analyses were run on the dataset with only completed cases. They show that  between past year identity theft was dependent on most of the predictors (p < 0.05) with the exception of sex (p > 0.05).

```{r chi, echo=FALSE}
chisq.test(its1$idtheft,its1$incomer)
chisq.test(its1$idtheft,its1$ager)
chisq.test(its1$idtheft,its1$sexr)
chisq.test(its1$idtheft,its1$ethnicr)
chisq.test(its1$idtheft,its1$prevent_total)
chisq.test(its1$idtheft,its1$OUTSIDE_PAST_YEARR)
chisq.test(its1$idtheft,its1$notify_breachr)
```

## Machine Learning
## Setup
### The dataset was split into training and testing datasets to in an attempt to train a logistics regression model predicting past year identity theft. Sixty percent of data was assigned to the training data while the remaining 40% went to the test dataset. The training dataset was examined for variables that had near zero or zero variability. There were no variables with zero or near zero variability.

```{r mlsetup, echo=FALSE}
its1$ager<-relevel(its1$ager, ref = 4)
its1$ethnicr<-relevel(its1$ethnicr, ref = 2)
its1$notify_breachr<-relevel(its1$notify_breachr, ref =2)

inTrain<-createDataPartition(its1$idtheft,p=.6, list=FALSE)
training <- its1[inTrain,]
testing <- its1[-inTrain,]
nzv<-nearZeroVar(training, saveMetrics = TRUE)
nzv
```

## Training Model
### A logistic regression model predicting past year identity theft was trained to the data. All of the predictors with the exception of sex was included in the model. Sex was excluded due to the chi-square analysis showing that sex was independent of past year identity theft. The results of the logistic regression model showed that all predictors were statistically significant in predicting past year identity theft (p<.05). Based on the odds ratios, 

```{r logistic}
modFit <- glm(idtheft ~incomer+ager+ethnicr+prevent_total+
                OUTSIDE_PAST_YEARR+notify_breachr,data=training,family="binomial")
#Training model estimates
summary(modFit)
#Odds ratios
exp(modFit$coeff)
```

## Training Model Diagnostics

```{r diagnostics}
#graph of model residuals
plot(modFit)
#get variance inflation factors
vif(modFit)
#anova- used for putting factors in and out of model
anova(modFit, test = "Chisq")
```

## Running Adjusted Model